# Data Cleaning {#chapter-2}

```{r, echo = F}
button <-  "position: relative; 
            top: -25px; 
            left: 85%;   
            color: white;
            font-weight: bold;
            background: #4B9CD3;
            border: 1px #3079ED solid;
            box-shadow: inset 0 1px 0 #80B0FB"
```

```{r, echo=FALSE, results='asis'}
codefolder::bookdown(init = "show", style = button)
```

In Chapter 2 we will work through some basic data cleaning operations useful in longitudinal data analysis. The basic idea is provide a set of scripts to use for exploring new repeated measures data sets.

## Example Data 

For Chapter 2 we will make use of the longitudinal Wechsler Intelligence Scale for Children [WISC; Wechsler, -@wechsler1949] dataset described by @osborne1972. These data have been detailed extensively in a number of papers [@mcardle1987; @mcardle1988; @mcardle1990; @mcardle1994] and are used here with with permission. 

The WISC data contains repeated measures data from 204 children between the ages of 6 and 11 years old (during grades 6, 7, 9 and 11). Thee repeated measures include component scores for the verbal tests and performance subtests at all four occasions, along with verbal subtest scores for the information, comprehension, similarities, and vocabulary domains at the first and last measurement occasion. The demographics variables mother’s education (continuous in years) and mother graduated high school (dichotomous) are also included.

## Reading in Repeated Measures Data

We can read in the WISC data directly from the [QuantDev](https://quantdev.ssri.psu.edu/) website.

```{r}
filepath <- "https://quantdev.ssri.psu.edu/sites/qdev/files/wisc3raw.csv"
wisc3raw <- read.csv(file=url(filepath), header=TRUE)
```

Additional details on importing different data types into R can be found here: http://www.statmethods.net/input/importingdata.html.

## Familiarize Yourself with the Data

Let's take an initial look at the structure of our data object using `str()`

```{r}
str(wisc3raw)
```

From the output, we can also see that the data frame consists of 204 observations (rows) and 20 variables (columns). Each variable’s name and data type is also listed. Methods like the ones above can be an effective way to initially familiarize yourself with the main features of a dataset. 


## Look for Duplicated IDs

It is always worth looking for non-unique ID numbers when  ID labels are included in a dataset. Here we have an `id` variable indicating the subject number. Since our data is in a long format (more on that later) duplicate IDs may indicate a potential problem with the data source or clues on how the data is structured. 

```{r}
any(duplicated(wisc3raw$id))
```

## Using `table()` to Spot Irregularities

When a variable takes on a limited range of values it is often useful to screen for irregularities or invalid values. This is common across all variable types and can occur for character strings, numeric, integer and factor types. For example, we would expect the `grad` variable to only take the values of zero or one. We can use the `table()` function to quickly confirm this. 

By default `table()` simply omits any values coded as `NA`. To include a count of the `NA` values use the `useNA` argument of `table()` as follows:

```{r}
table(wisc3raw$grad, useNA = "always")
```

## Missing Data 

Dealing with missing data in a consistent manner is one of the most important aspects of data cleaning. When data are imported into R it is common to discover missing values are coded according to a variety of conventions. 

Often a first step in handling missing data involves recoding missing values as `NA`. Writing bespoke code to handle the different types of missing data one might encounter is tedious and unnecessary. 

`naniar` [@naniar] is a useful package with many convenience functions for managing missing data in R. Here we demonstrate some of this functionality.

### Generating Example Data

Since the WISC data does not contain missing values it is helpful to generate a synthetic dataset containing some commonly encountered missing data codes. 

```{r}
set.seed(123)
wisc_miss <- wisc3raw
wisc_miss$verb1[sample(nrow(wisc_miss),100)] <- -99
wisc_miss$comp1[sample(nrow(wisc_miss),75)] <- "N/A"
wisc_miss$info1[sample(nrow(wisc_miss),50)] <- "NA"
```


### Recoding Values with `NA`

Now that we have a dataset with missing values we can use `naniar` to recode these values to `NA`.


```{r}
na_strings <- c("NA", "N/A", -99)
  
wisc_miss <- naniar::replace_with_na_all(
  wisc_miss, condition = ~.x %in% na_strings
)
```

See the [`naniar` vignette on recoding NA values](https://cran.r-project.org/web/packages/naniar/vignettes/replace-with-na.html) for more detailed information on the package functionality.

### Missing Data Visualization

Once we have recoded our data in a consistent manner we can use visualizations to explore the missing data. The `vis_miss()` function from `naniar` is a good starting point for visualizing the amount of missing data in our dataset. The plots shows the missing values in black and non-missing values in gray. In addition, percentages of missing data in both the dataset and individual variables are provided.

```{r}
naniar::vis_miss(wisc_miss)
```

Many missing data visualizations are described in the [`naniar` vignette on missing data visualization](https://cran.r-project.org/web/packages/naniar/vignettes/naniar-visualisation.html) including plots for exploring missing data mechanisms. 

## Exporting Data

Depending on work-flow, you may need to export your dataset for use in another statistical software program. The `write.csv()` function is a convenient method for outputting comma delimited files.

```{r}
write.csv(wisc3raw, file = "wisc3raw.csv", row.names = FALSE, na = "-99")
```

Note that by default the `write.csv()` function will include an extra column of row numbers and will notate missing data with an `NA`.  More information on exporting data is available at http://www.statmethods.net/input/exportingdata.html.


## Reshaping Repeated Measures Data 

Behavioral science tends to use relational data structures - in basic form, spreadsheets. Typically, the data are stored  in a data frame (a "fancy" matrix) with multiple rows and columns. Two common schemata used to accommodate repeated measures data are *wide format* and *long format*. Different analysis and plotting functions require different kinds of data input. Thus, it is imperative that one can convert the data back and forth between wide and long formats. 

There are lots of ways to do this. We illustrate one way.

*Sidebar*: The `dput()` function provides a convenient method to get the variable names (or any R object) into a format that can be read back into R. For example, this can be helpful when working with a long vector of strings.

```{r}
dput(colnames(wisc3raw))
```

First, let's subset our data to only include the variables we need for this analysis.

```{r}
var_names_sub <- c(
  "id", "verb1", "verb2", "verb4", "verb6",
  "perfo1", "perfo2", "perfo4", "perfo6",
  "momed", "grad"
)

wiscraw <- wisc3raw[,var_names_sub]
head(wiscraw)
```


### Reshape Wide to Long

One way to go *from wide to long* is using the `reshape()` function from base R. 

Notice, the `varying` argument contains the repeated measures columns we want to stack and the `timevar` is a new variable containing the grade level information previosuly appended at the end of the colnames listed in `varying`.

```{r}
# reshape data from wide to long
wisclong <- reshape(
  data = wiscraw,
  varying = c("verb1", "verb2", "verb4","verb6", "perfo1","perfo2","perfo4","perfo6"),
  timevar = c("grade"), 
  idvar = c("id"),
  direction = "long", 
  sep = ""
)

# reorder by id and day 
wisclong <- wisclong[ order(wisclong$id, wisclong$grade), ]

head(wisclong, 8)
```  

Again, notice how `reshape` automatically split verb1, verb2, etc. into a string name and a grade variable. 

### Reshape Long to Wide

Now we go *from long to wide*, again using the `reshape()` function. The `v.names` argument specifies the variables to be expanded column wise based on the repeated measure specified in `timevar`.

```{r}
#reshaping long to wide
wiscwide <- reshape(
  data = wisclong, 
  timevar = c("grade"), 
  idvar = c("id"),
  v.names = c("verb","perfo"),
  direction = "wide", 
  sep = ""
)

# reordering columns 
wiscwide <- wiscwide[, c(
  "id", "verb1", "verb2", "verb4", "verb6",
  "perfo1", "perfo2", "perfo4", "perfo6",
  "momed","grad" 
)]

head(wiscwide)
```                    

Using functions included in base R can be useful in a number of situations. One example is package development where one may wants to limit dependencies. 

That said, many people find `reshape` to be unnecessarily complicated. A similar, and potentially more convenient, set of functions have been developed for reshaping data in the `tidyr` [@tidyr] package. For those interested take a look at the `pivot_longer()` and `pivot_wider()` functions. 

For examples using `tidyr` to reshape data see the [tidyr vignette on pivoting](https://cran.r-project.org/web/packages/tidyr/vignettes/pivot.html).
